```{r echo=FALSE, include=FALSE} 
library(tidyverse)
```

```{r, echo=F, message=F}
stop_features = read_csv('data/stop_features.csv')
```



# Background

Public transit is a very popular subject to research across a broad range of topics, including: investigating effects of COVID on public transit demand [@covid], effects of public transit on physical activity [@activity], effects on ridesharing and biking [@bikes], effects on employment [@jobs], equity-advancing practices [@equity], and more. 
However, this study focuses specifically on TriMet and the qualities and features of TriMet station stops. Therefore, any similar previous studies have been undertaken by TriMet themselves. TriMet has a large variety of data-driven infrastructure and safety projects in the works, a list of which can be found at [@trimet_projects], but the data and analysis that drive these projects are kept internally. TriMet's commitment to transparency extends to making their data available, with an entire page on their website dedicated to transparency [@transparency], but not any studies done. In this analysis, we provide a publicly available inquiry into TriMet’s data.

# Ethical Considerations 

Access to public transit can lead to difficult ethical considerations. Neighborhoods with historically underprivileged populations are often the ones that rely most on public transit, but those neighborhoods are also the ones that are least likely to receive transit stations within a usable distance. 
The idea of what counts as a "usable distance" is itself a difficult question to answer ethically. This can even vary for the same person over time–walking for 10 minutes on the way to work might be a nice part of the commute, even in the rain, but becomes a slog when hauling luggage to the airport. A distance that takes one person ten minutes to walk may take someone else a half hour, and they have no energy to do errands once they've arrived at the bus stop. These are not easy considerations to make, and yet they are crucial ones. Transit centers can get around this issue somewhat by providing parking, but that still requires someone to have a car in order to get to the transit center to take a transit route–or requires having a larger number of local transit routes that feed into the transit center from the surrounding area. 
Designing and running an effective transit system requires grappling with ethics to ensure it is serving the most people possible in the most effective ways. 

All data used in this project is publicly available and does not contain information about protected classes, so it requires no ethical considerations when being worked with. 
