```{r echo=FALSE, include=FALSE} 
library(tidyverse)
```

```{r, echo=F, message=F}
stop_features = read_csv('data/stop_features.csv')
```



# Background

Recent studies specific to Portland public transit include investigating effects of COVID on public transit demand, effects of public transit on physical activity, effects on ridesharing and biking, effects on employment, equity-advancing practices, and more. However, this study focuses specifically on TriMet and the qualities and features of TriMet stops. Therefore, any similar previous studies have likely been done by TriMet themselves. Our research is certainly not more in-depth than TriMet’s, since they have easier access to their own data. We hope to provide a publicly available inquiry into TriMet’s data.

# Ethical Considerations 

Access to public transit can lead to difficult ethical considerations. Neighborhoods with historically underprivileged populations are often the ones that rely most on public transit, but are also those that are least likely to receive transit stations within a usable distance. 
The idea of what counts as a "usable distance" is itself a difficult question to answer ethically. This can even vary for the same person over time–walking for 10 minutes on the way to work might be a nice part of the commute, even in the rain, but becomes a slog when hauling luggage to the airport. A distance that takes one person ten minutes to walk may take someone else a half hour, and they have no energy to do errands once they've arrived at the bus stop. These are not easy considerations to make, and yet they are crucial ones. Transit centers can get around this issue somewhat by providing parking, but that still requires someone to have a car in order to get to the transit center to take a transit route–or requires having a larger number of local transit routes that feed into the transit center from the surrounding area. 
These are only a few of the ethical considerations that must go into designing and running a transit system. 

All data used in this project is publicly available and does not contain information about protected classes, so it requires no ethical considerations when being worked with. 
